---
title: "Data Manipulation Assignment - Rachael Lehner"
output: html_document
date: "2025-10-09"
---

#Part 1
1.Open the sleep dataset in R 
```{r}
data(sleep)
```

2. Is it long or wide format? How do you know? 
*This is long data, because if it were wide data, the groups would have their own columns. Since the groups are repeated multiple times in the group column, I think it's long data.*

3. Transform this data so that it is in the other format. 
```{r}
sleepWide<- pivot_wider(sleep, id_cols = "ID", names_from = "group", values_from = "extra")
```

4. Transform your data back to its original format. 
```{r}
sleepLong<- pivot_longer(sleepWide, cols = 2:3, names_to = "group", values_to = "extra")
```

5. Open the Starwars dataset
```{r}
data("starwars")
```

6. Create homeworld as a factor. How many levels does it have? 

```{r}
homeworldFactor<- as.factor(starwars$homeworld)
#This factor has 48 levels 
```

7. Reorder homeworld so that Naboo is the reference group. 
```{r}
relevel(homeworldFactor, ref = "Naboo")
```

8. Use mutate to create a column transforming kg to pounds. (hint: you may need to look up the conversion)
```{r}
newStarswarsDat<- starwars %>% mutate(pounds = mass*2.2)
```

9. Print a summary of your column from #8
```{r}
summary(newStarswarsDat$mass)
```

#Part 2
10. open the billboard dataset.
```{r}
data("billboard")
```
11. are the data in this dataframe in long or wide format? explain how you know. 
*I think the data is wide because the weeks all have their own column. If it were long data, I think there would just be one big column for all the weeks and the numbered weeks would just appear repeatedly.*

12. change the data from its original format to the other version. 
```{r}
longBillboard<- pivot_longer(billboard, cols = 4:79, names_to = "weeks", values_to = "score")
```

13. change it from your altered version in #3 back to its original state. 
```{r}
wideBillboard<- pivot_wider(longBillboard, id_cols = c("artist", "track", "date.entered"), names_from ="weeks", values_from = "score")
```

14. Separate the date entered column into three columns: one for year, one for month, and one for day.

```{r}
billboardSepareated<- separate(billboard, date.entered, into = c("year", "month"), sep = 5)
billboradSepFinal<- separate(billboardSepareated, month, into = c("month", "day"), sep = 3)
billboradSepFinal
```

