---
title: "Ggplot Assignment"
output: html_document
date: "2025-10-27"
---
---
title: "Ggplot part 1 assignment"
author: "Amanda Mae Woodward"
date: "2025-10-23"
output: html_document
---

1. open the starwars dataset in tidyverse
```{r}
library(tidyverse)
data(starwars)
```


2. create a graph showing the frequency of eye colors in the dataset. 
```{r}
ggplot(starwars, aes(eye_color))+geom_bar()
```


3. Customize your graph to make it more presentable. Consider color, backgrounds, axes, etc. 
```{r}
ggplot(starwars, aes(eye_color))+
  geom_bar(color ="#660033", fill= "#ffe6ff")+
  theme_classic()+ 
  xlab("Eye Color")+ 
  ylab("Frequency")+
  ggtitle("Frequency of Eye Color in Star Wars Characters")
```


4. Create a graph displaying information about character's height in each homeworld. 
```{r}
ggplot(starwars, aes(height, fill = homeworld))+geom_histogram()
```


5. Customize your graph to make it more presentable. Consider color, backgrounds, axes, etc.
```{r}
library(viridis)
```

```{r}
ggplot(starwars, aes(height, fill = homeworld))+
  geom_histogram(bins = 15)+
  xlab("Height (cm)")+
  ylab("Frequency")+
  ggtitle("Starwars Characters' Height By Homeworld")
  theme_classic()
```


6. Create a graph of character's height by mass. 
```{r}
ggplot(starwars, aes(height, mass))+geom_point()
```


7.Customize your graph to make it more presentable. Consider color, backgrounds, axes, etc.
```{r}
ggplot(starwars, aes(height, mass))+
  geom_point(color="#660033", size =2, shape = 16)+
  geom_smooth(method = "lm", size = .5, color = "black", linetype = "dashed")+
  theme_classic()+
  xlab("Height (mm)")+
  ylab("Mass (kg)")+
  ggtitle("Star Wars Characters' Height by Mass")
```


8. Add gender as a third variable to your graph of height by mass. 
```{r}
ggplot(starwars, aes(height, mass, color = gender))+geom_point()
```


9. Customize your graph to make it more presentable. Consider color, backgrounds, axes, etc.
```{r}
ggplot(starwars, aes(height, mass, color = gender))+
  geom_point(size = 3)+
  geom_smooth(method = "lm", size = .5, color = "black", linetype = "dashed")+
  theme_classic()+
  xlab("Height (mm)")+
  ylab("Mass (kg)")+
  ggtitle("Star Wars Character's Height by Mass and Gender")
```

