---
title: "PSY 4802 Notes (2025-09-22)"
output: html_document
date: "2025-09-23"
---
---
title: "4802 Notes (2025-09-16)"
output: html_document
date: "2025-09-16"
editor_options: 
  markdown: 
    wrap: 72
---

---
title: "Foundations in R part 1"
author: "Amanda Mae Woodward"
date: "2025-09-16"
output: html_document
---

# Learning Outcomes:

After Today's Lecture, you should be able to do the following: -
Evaluate mathematical expressions and logical statements in R - Describe
how to use common R functions - Explain the differences between
variables, vectors, and dataframes - Explain how to obtain a subset of
your data - Describe your data effectively

# Panels of R Studio

## top left: R Markdown

-   where we put stuff we want to save

## top right global environment:

-   has everything (overview)
-   contains variables, datasets, vectors, functions

## bottom left: console

-   where code runs (see output here depending on settings) - errors
-   scratch paper (doesnt save)
-   `install.packages()` & `View()`

## bottom right:everything else

**Disclaimer:** I am not a computer programmer. I occasionally use
programming words and may use them incorrectly (feel free to correct
me!). I prefer to use words like "thing" and "stuff." If it gets
confusing, just let me know.

### code chunks

text goes outside

```{r}
# code -> insert chunk
# Mac: command + option + i --> I like to use this one
# pc: control + alt+i
# manual
```

## Learning Outcome 1: Mathematical Expressions and Logical Statements

### Math

R can be treated like a big fancy calculator (think scientific
calculators you may have used previously), but can do much more.

#### Addition

```{r}
#code = number + number (command + return on mac runs the code)
4+9
```

#### Subtraction

```{r}
#code = number - number (command + return)
7-5
```

#### Multiplication

```{r}
#code = number*number
426*315
```

#### Division

```{r}
#code = number/number
67435/162
```

#### Square Root

`sqrt()`

```{r}
#code = sqrt(number)
sqrt(27463)
```

#### Exponent

```{r}
#code = number^number
452^4
```

**Important:** In terms of calculations, R follows PEMDAS (algebra
throwback). Parentheses are going to be **Really** important.

#### More complicated Math:
*uses PEMDAS*

```{r}
(3+5)^6
3+5*6
```

### Logical Statements

We're going to go through logical statements because they can be helpful
when wrangling/cleaning data.

First, what's a logical statement? It evaluates whether an expression is
TRUE or FALSE

Consider the following sentence:

"Michael Scott on *the Office* is a good manager"

If this sentence was a fact, R would output the word "TRUE." If it was
not true, R would output the word "FALSE."

To assess whether logical statements in R are true, we use two equal
signs `==`. The above example would look something like this:



```{r}
#code = "thing1" == "thing 2" --> is value on the right the same as the value on the left
"michael scott" == "good manager"
```

Running this code tells us the statement is false. (**Note**: this is
because the phrase on the left is not the same as the one on the
right..as far as I know, R hasn't watched *the Office*).

```{r}
#code= "thing1" == "thing2" OR number == number 

3 == 4
2*6 == 3*4

```

However, R can compute mathematical expressions. This means we can look
at something like this:

```{r}
3 == 4
2*6 == 3*4
```

You can also evaluate if statements are **NOT** True. We can do this by
using the symbol `!=` For instance, I could evaluate Michael Scott as a
manager by writing:

```{r}
#code = "thing1" != "thing2" OR number != number

"michael scott" != "good manager"

```

Note: In other words, == is similar to "is" and != is similar to "is
not"

A similar process can be used to determine if something is greater than,
less than, greater than or equal to, and less than or equal to.
**Aside:** I use these a lot to clean developmental data

Examples:

```{r}
#code = number< number (For greater than, just swap the "<" for ">")
54< 26
54< 87

#less than or equal to code = number <= number (For greater than or equal to, just swap the "<" for ">")
54<=67
54<=34
```

We can also make logical statements more complex by creating **and** and
**or** statements.(And- both things need to be true Or- only one has to
be true)

And statements use `&` OR `&&` (if you have more things to evaluate)

```{r}
#code = number sign number & number sign number 

3+5 == 8 & 32>41
3+5 ==8 && 32>41 && 21<67


```

Or statements use `|` OR `||` (if you have more things to evaluate)

```{r}
#code= number sign number | number sign number 

3+4==8 | 32<42
32<33 || 3-6==2 || 21-7==14
```

##### Learning Outcome 1 Practice:

1.  Find the square root of 89

```{r}
sqrt(89)
```

2.  Determine if 3 raised to the 5 power is greater than the square root
    of ten thousand

```{r}
3^5>sqrt(10000)
```

3.  Determine if 5 times 7 is less than or equal to 30 or if 4\*8 is
    equal to 32

```{r}
5*7<=30 | 4*8==32
```

### Learning Outcome 2: Describe how to use common R functions

A note on R functions: R has built in functions that help us create,
manipulate, and transform our data. Generally, we use functions by
writing the function name followed by parentheses. We'll talk about more
of them throughout this lesson, and during the semester. Here are some
that are just generally helpful.

If you want to learn more about a function, you can type `?` function
name in the console.

The first one we'll talk about is `:`. Typically, it would be used by
writing small#:larger#, and would give you all of the numbers in
between.

```{r}
#code = small number : large nunmber --> looks at all numbers from a:b

1:100
100:1
```

`rep()` allows you to repeat something a set number of times. Looking at
the help file, we can see the arguments contained in the function. To
use it, we write `rep(thingToBeRepeated, #ofTimes)`.

```{r}
#code rep(number you want to repeat, times you want to repeat it) OR rep("word you want to repeat", times you want to repeat it)

rep(15, 7)
rep("salt", 5)
```

**Note:** For functions, you can either write arguments in the order
listed in the help file (not labelled), or make sure you include the
argument label.

```{r}
#rep(number OR "word", times = the amount of times you want to repeat)
rep(x=12, times = 2)
rep(x="bag", times = 4)
```

`seq()` gives you a sequence of numbers. typically, we use it by writing
`seq(lowest#, highest#, by=#)`.

```{r}
#seq(lowest#, highest#, by=#) --> don't have to write "by, could just use a comma 
```

#### Learning Outcome 2 Practice:

1.  Print every 5th number between 1 and 100.

```{r}
seq(1,100,5)
```

2.  Use code to write I love statistics 4 times.

```{r}
rep("I love statistics", 4)
```

3.  Repeat the numbers 1-10 twice

```{r}
rep(seq(1,10),2)
```

4.  Print every 3rd number from 1-20 3 times.

```{r}
rep(seq(1,20,3),3)
```

###Learning Outcome 3: Explain the differences between variables,
vectors, and dataframes

What's the problem with using R as a big, fancy calculator? We haven't
stored anything so we could use it later. Different things that we will
need later have different names, but we store them in the same way

nameItsBeingSavedAs \<- thingBeingSaved

nameThing = thingSaved #DO NOT DO THIS

**Note:** the "assignment operator" (less than sign followed by the
hyphen; I was taught to call it a carrot) should be used over an equal
sign. There are a lot of reasons for this, but the main one is that R is
finicky and different updates and packages read "=" differently, so \<-
is safest.

Naming conventions: I was taught to use all lower case for the first
word, and capital letters for each word after. That's what I'll use in
class. However, you can use whatever convention you want. But a couple
of things to keep in mind:

-   R objects (things) can't have spaces in the name
-   R is case sensitive (statistics, Statistics, StAtIsTiCs are all read
    as three different things)
-   Don't give your stored stuff names that are the same as R functions
    (e.g. no naming dataframes "data" or saving something as "mean")
-   Make sure your names mean something (slgahwgh may help you get our
    your frustration out- but you won't remember what it means later and
    neither will your collaborators)

**Variables**

Variables store one thing in R (it can be a number, a word, or whatever
you want)

```{r}
#code = variableName <- thingBeingSaved
# DO NOT DO: VariableName = thingBeingSaved
friut <- "blurberries"
coffeeConsumes <- 0
```

\*Aside: `c()` means combine. you can put stuff in parentheses and use a
comma to separate it and have it print together

**Vectors**

Vectors are objects that store one dimensional information

To save more than one thing in an R variable, you would use the combine
function- `c()`.

```{r}
"numeric"
#code =  vectorName<- c(data value, data value, data value, ect.)
darkMatter<- c(0,0,0,1,0,4,3)
"character"
#cosde = vectorName<- c("data point", "data point", "data point", etc.)
firstNames<- c("Rachael", "Tori", "Sam", "Amanda")
```

If you want to call one of the things you saved in a vector, you can
write `vectorName[position of thing]`

```{r}
#cade = vectorName[position of thing you want]
firstNames[3]
```

**Dataframes** Data frames are made from multiple vectors. There are
preloaded datasets in R we can use to practice, such as mtcars:

```{r}
#code = data(mtcars)
data(mtcars)
```

**2025-09-18**
### Practice:

1)Make a vector with 6 animals

```{r}
Animals<- c("Jaguar", "Quetzal", "Turkey", "Wolf", "Chinchilla", "Armadillo")
```

2)Make a vector of the number of cookies you think each animal could eat

```{r}
animalCookies<- c(20,3,9,18,1,7)
```

We can also combine vectors into a dataframe using R functions
(`cbind.data.frame()` or `rbind.data.frame()`). We'll use the vectors we
made above as an example.

```{r}
# code = dataName<- cbind.data.frame(vector1, vector2) -- puts things in columns 
cookieData<- cbind.data.frame(Animals, animalCookies)

#code = dataName<- rbind.data.frame(vector1, vector2) -- puts things in rows
cookie2Data<- rbind.data.frame(Animals, animalCookies)
```

Just like you can pick one thing in a vector, you can select one column
or one row in a dataframe. You can use brackets like we did with
vectors, but you need to specify whether it's a row or a column. If it's
a column, you can write `dataframe[,column#]` and you can use
`dataframe[row#,]`.

```{r}
cookieData[,1] #gives you the first columns
cookieData[1,] #gives you the first row
cookieData[5,2] # gives you a single cell (x position in row, y position in column)

```

You can also get columns by using a `$`.

```{r}
#code = dataName$Variable of interst
cookieData$Animals

```

#### Learning Outcome 3 Practice:

1.  Create a variable that contains your favorite ice cream flavor

```{r}
favIceCream <- "Mint Chocolate Chip"
```

2.  Create two vectors: 1 vector should contain the names of four TV
    shows and the 2nd vector should contain your ratings for those shows

```{r}
favTVShows<- c("Attack on Titan", "Jujutsu Kaisen", "Modern Family", "Stranger Things")
showRatings<- c(5,5,5,5)
```

3.  Make these vectors into a data frame.

```{r}
TVdata<- cbind.data.frame(favTVShows, showRatings)
```

4.  Print the information in the 2nd row.

```{r}
TVdata[2,]
```

5.  Print the information in the 2nd column.

```{r}
TVdata[,2]
```

Bonus: Print the information in the 3rd row, 1st column.

```{r}
TVdata[3,2]
```

###Learning Outcome 4: Explain how to obtain a subset of your data

Sometimes, you don't need a full dataframe, you just want to use a piece
of it. To do this, we can use subsetting or indexing to save just that
piece.

**Indexing** We'll started with indexing because it is similar to what
we did above. You can write dataframeName[logical statement] to get a
portion of the data.

For instance, we could use indexing to get only cars in the mtcars
dataset with 4 gears.

```{r}
#code = dataset[dataset$variable == thing of interest,] -- use the , for the column info too
mtcars[mtcars$gear==4,]

#make a new data set
#code = new dataset <- dataset[dataset$variable == thing of interest,]
fourGears<- mtcars[mtcars$gear==4,]
```

In this case, you'll notice that I put a comma after the logical
statement. This is because we want all of the **rows** where the car has
4 gears.

**Subsetting** We can also use the subsetting function to create a new
dataframe. We can look at the subset function by typing ?subset in the
console. We use this function by `subset(dataframe, logical statement)`

If we wanted to get the same subset of data as above, we could do the
following:`subset(dataframe, condition)`

```{r}
#five gears and 6 cyl
#code = dataName<- subset(data, data$variable== What you want the vairble limit to be) #Note: you can use other logical fundtion like greater than or equal to (>=) as well as add additional varibles using the & sign
fiveGearSixCyl<- subset(mtcars, mtcars$gear== 5 & mtcars$cyl==6)
fiveGearSixCyl
```

**Subsetting Practice**
```{r}
TwentyFiveMps<- subset(mtcars, mtcars$mpg < 25)
AutoTrans<- subset(mtcars, mtcars$am<1)
oneHundredHpFourGr<- subset(mtcars, mtcars$hp>=100 & mtcars$gear==4)
```

can combine logical statements in all the ways we talked about before

**A Note on Pipes (more advanced)** If not, we're going to talk about
another way to get a subset of data from our dataframe.

Above, we used indexing and subsetting to make new dataframes that
contain only the information we need. You can also use something called
pipes to create a subset (For those who use tidyverse). A pipe, or `%>%`
is another way to program in R. It takes whatever is on the left and
"feeds" it to the function on the right. So the data goes through a
pipeline to get you the desired outcome.

First, we need to load a package. Package libraries have lots of
functions made by really smart people to make our lives easier. To load
a package, you need to first use the following code in the console:
`install.packages("packageName")`

After that, we need to use the library command to load the package

```{r}
#code = install.packages("Name of Package you want")
#install.packages("tidyverse")
library(tidyverse)
```


**2025-09-23**
We can use filter and select to subset data with pipes.

We can use `filter()` to pick a subset of rows:

```{r}
#code = subsetName <- dataset %>% filter(variableOfInterest EquationSign VariableLimit)
#Note: can do the %>% by comand+shit+m
gears<- mtcars %>% filter(gear==5)
```

We can use `select()` to pick a subset of columns:

```{r}
#code = subsetName<- dataset %>% select(varibleOfInterst) #Note: can do many varibles inside parenthasis, just use a comma to separate
MpgHp<- mtcars %>% select(mpg, hp)
```

```{r}
#Note: can combine filter and select
#code:
#subset<- dataset %>% 
  #filter(VariableOfInterst EquationSign VariableLimit)
  #Select(VariableOfInterst)

MpgGears<- mtcars %>% 
  filter(gear==5) %>% 
  select(mpg, hp)
```

#### Learning Outcome 4 Practice:

1.  Load the ChickWeight dataset.

```{r}
data("ChickWeight")
```

2.  Create a subset of data that contains information for Chick 8

```{r}
ChickEight<- ChickWeight %>% filter(Chick==8)
```

3.  Create a subset of data for Diet 2 only.

```{r}
DietTwo<- ChickWeight %>% filter(Diet==2)
```

4.  Create a subset of data for all chicks weighing less than 100 gm.

```{r}
SmallChicks<- ChickWeight %>% filter(weight<100)
```

### Learning Outcome 5: Describe your data effectively

We need to make sure that we can understand what our data look like
before doing anything more complicated.

**Looking at the top/bottom of your dataset** First we can view what our
data look like using `head()` and `tail()` to see the beginning and end
of our data. they both use a similar format
`head(dataframe, # of rows )` We'll use the diamonds dataframe (need to
install ggplot if you haven't already!)

```{r}
#code = head(dataset) -- gives first six rows
#code = heat(dataset, RowNumberOfInterest) -- shows you the rows you want to see

head(mtcars)
head(mtcars,3)
```


```{r}
#code = tail(dataset) -- gives you bottom six rows
#code = tail(dataset, RowNumberOfInterest) -- shows you the rows you want to see

tail(mtcars)
tail(mtcars, 3)
```

This lets us see just part of our data to make sure it loaded
appropriately/is what we expect.

**Summary** We can also describe our data using a function called
`summary()`.

```{r}
#code = summary(dataset)
#can also be used on a single variable 
#code = summary(dataset$VariableOfInterst)
summary(mtcars)
summary(mtcars$mpg)
```

What do you notice about the summaries?

**Aside:** There are different classes(types) of data. They get
summarized differently, and the class of data influences what functions
you can run.

There are **many** functions to summarize data, and we'll use more as
the semester goes on.

**Descriptive Statistics** We can also get descriptive statistics for
specific columns

If we need the \# of observations, we can use `length()` or `nrow()` to
get this information.

```{r}
#code length(dataset) -- works on vectors and single variables 
#code length(dataset$variableOfInterst)
#nrow(dataset) -- tells you number of rows 
#ncol(dataset) -- tells you number of columns 

length(mtcars$mpg)
length(mtcars)
nrow(mtcars)
ncol(mtcars)
```

We can also take the average by using the function `mean()`.

```{r}
#code = mean(dataset$Varible) -- gives the mean 
mean(mtcars$mpg)
```

We can get the standard deviation using `sd()`.

```{r}
#code - sd(dataset#varble) -- gives typical distance from the mean 
sd(mtcars$mpg)
```

`var()`gives us the variance

```{r}
#code var(dataset$Variable) -- gives variance (sd squared I think)
var(mtcars$mpg)
```

`range()` gives us the range

```{r}
#code = range(dataset$Variabll) -- Gives high and low value, subtract them for the range 
range(mtcars$mpg)
33.9-10.4
```

###Learning Outcome 5 Practice

1.  Summarize the ChickWeight dataset

```{r}
summary(ChickWeight)
```

2.  Find the mean of the weight column

```{r}
mean(ChickWeight$weight)
```

3.  Summarize the data for just the 1st chick

```{r}
ChickOne<- subset(ChickWeight, Chick==1)
summary(ChickOne)
#OR
summary(ChickWeight %>% filter(Chick==1))

```

4.  What is the variance of the 1st chick's weight?

```{r}
var(ChickOne)
```

5.  What is the range in weight for each chick posted in the dataset?

```{r}
range(ChickWeight$weight[ChickWeight$Chick==1]) #Then copy this down but change the number 
range(ChickWeight$weight[ChickWeight$Chick==2])
```