---
title: "Quiz 1 - Rachael Lehner"
output: html_document
date: "2025-10-09"
---


This quiz will cover everything we've covered in foundations of R assignments.

1. Open the tidyverse library.
```{r}
library(tidyverse)
```

2. Open the starwars dataset.
```{r}
data("starwars")
```

3. Create a subset of data that contains only the humans from the starwars franchise.
```{r}
starwarsHumans<- subset(starwars, starwars$species=="Human")

```

4. In this subset of data, who is the tallest character? 
*Darth Vader -- I just looked at the data, I didn't need to run a code*


5. In this subset of data, print the information about the character in the third row. 
```{r}
characterThree<- subset(starwars, starwars$name == "R2-D2")

```
```{r}
characterThree
```

6. In the full starwars dataset, calculate the average mass for the masculine characters. 

```{r}
mascChar<- subset(starwars, starwars$gender=="masculine") #use gender 
mean(mascChar$mass, na.rm = TRUE) 
```

7. Create a dataset that contains only the characters and the films that they appeared in from the full dataset. 
```{r}
filmsAndChars<- starwars %>% 
  select(name, films)
```

8. Create a new column in the starwars dataset that contains a "ranking" of all characters (Note: you don't have to actually rank them. Just fill the column in with the numbers 1-87)
```{r}
rank<- sample(size = 87, 1:87, replace = TRUE)
CharsRate<- data.frame(name = filmsAndChars$name, rank)
filmsCharsRate<- left_join(filmsAndChars, CharsRate)
filmsCharsRate
```

9. Create a subset of data for the characters from Naboo. How many characters are from Naboo? 
```{r}
Naboo<- subset(starwars, starwars$homeworld == "Naboo")
Naboo
#There are 11 characters from Naboo
```

10. How many droids are in the full dataset? 
```{r}
droids<- subset(starwars, starwars$species == "Droid")
droids
#6 characters are droids
```

